---
title: "Example_time_series_example_comparison"
output: html_document
---

The goal of this section is to compare the results from edge (df = 2) and maSigPro (df = 2) for the endotoxin data.

```{r}
# Load libraries

library("Hmisc")
library("formattable")

# Load data

# p-values from edge

p_value_endo_genes_edge_data <- read.csv("~/diffTimeExpression/data/p_value_endo_genes_edge_data.txt", sep="")

# p-values from maSigPro (df = 2)

p_value_maSigPro_df_2 <- read.csv("~/diffTimeExpression/data/p_value_maSigPro_df_2.txt", sep="")

# p-values from maSigPro (df = 5)

p_value_maSigPro_df_5 <- read.csv("~/diffTimeExpression/data/p_value_maSigPro_df_5.txt", sep="")

```


### Compare unadjusted p-values from edge and maSigPro (df= 2)

```{r}
# Combine the two lists of unadjusted p-values into one data frame
unadjust_p_value <- as.data.frame(cbind(p_value_endo_genes_edge_data, p_value_maSigPro_df_2))
colnames(unadjust_p_value) <- c("Unadjusted p-value in edge (df = 2)", "Unadjusted p-value in maSigPro (df = 2)")

# Find the correlation between the two numbers

rc <- rcorr(as.matrix(unadjust_p_value), type="pearson") # Correlation = 0.3766195 and p = approximately 0

flattenCorrMatrix <- function(cormat, pmat) {
  ut <- upper.tri(cormat)
  data.frame(
    row = rownames(cormat)[row(cormat)[ut]],
    column = rownames(cormat)[col(cormat)[ut]],
    cor  =(cormat)[ut],
    p = pmat[ut]
    )
}

flattenCorrMatrix(rc$r, rc$P)

# Compare unadjusted p-values from edge and maSigPro

plot(unadjust_p_value, main = "Unadjusted p-values from edge and maSigPro (Pearson's correlation = 0.377)")

# Make a best fit line (which we can then add to the plot)
abline(lm(unadjust_p_value[,1] ~ unadjust_p_value[,2]))

```

From this plot, it appears that edge is more conservative than maSigPro when df = 2.


### Compare unadjusted p-values from edge and maSigPro (df= 5)

```{r}
# Combine the two lists of unadjusted p-values into one data frame
unadjust_p_value <- as.data.frame(cbind(p_value_endo_genes_edge_data, p_value_maSigPro_df_5))
colnames(unadjust_p_value) <- c("Unadjusted p-values in edge (df = 2)", "Unadjusted p-values in maSigPro (df = 5)")

# Find the correlation between the two numbers

rc <- rcorr(as.matrix(unadjust_p_value), type="pearson") # Correlation = 0.3235 and p = 1.203482e-13

flattenCorrMatrix(rc$r, rc$P)

# Compare unadjusted p-values from edge and maSigPro

plot(unadjust_p_value, main = "Unadjusted p-values from edge and maSigPro (Pearson's correlation = 0.324)")

# Make a best fit line (which we can then add to the plot)
abline(lm(unadjust_p_value[,1] ~ unadjust_p_value[,2]))

```

The correlation for unadjusted p-values is higher in edge and maSigPro when df = 2 in maSigPro

### Compare Benjamini-Hotchberg corrected p-values from the two software programs

```{r}
# Make Benjamini-Hotchberg correction for each set of the unadjusted p-values

edge_fdr <- p.adjust(as.matrix(p_value_endo_genes_edge_data), method = "fdr", n = 500)
summary(edge_fdr)

maSigPro_fdr <- p.adjust(as.matrix(p_value_maSigPro_df_2), method = "fdr", n = 500)
summary(maSigPro_fdr)

# Combine the two lists of adjusted p-values into one data frame

adjusted_p_value <- as.data.frame(cbind(edge_fdr, maSigPro_fdr))
colnames(adjusted_p_value) <- c("BH adjusted p-values in edge (df = 2)", "BH adjusted p-values in maSigPro (df = 2)")

# Find the correlation between the two numbers

rc <- rcorr(as.matrix(adjusted_p_value), type="pearson") # Correlation = 0.4421 and p equals approximately 0

flattenCorrMatrix(rc$r, rc$P)

# Compare unadjusted p-values from edge and maSigPro

plot(adjusted_p_value, main = "BH adjusted p-values from edge and maSigPro (Pearson's corr. = 0.4421)")

# Make a best fit line (which we can then add to the plot)
abline(lm(adjusted_p_value[,1] ~ adjusted_p_value[,2]))

```

### Compare the number of genes at FDR 1% and FDR 10% for both software programs

```{r}
# Find the number of significant genes at FDR 1%

length(edge_fdr[edge_fdr<=0.01]) # 10 genes
length(maSigPro_fdr[maSigPro_fdr<=0.01]) #135 genes

# Find overlap

overlap_0.01 <- adjusted_p_value[adjusted_p_value[,1] <= 0.01 & adjusted_p_value[,2] <= 0.01, ]

nrow(overlap_0.01)

# Find the number of significant genes at FDR 10%

length(edge_fdr[edge_fdr<=0.1]) #61 genes
length(maSigPro_fdr[maSigPro_fdr<=0.1]) # 196 genes

overlap_0.1 <- adjusted_p_value[adjusted_p_value[,1] <= 0.1 & adjusted_p_value[,2] <= 0.1, ]

nrow(overlap_0.1)

# Make a table of the results

DF <- data.frame(Program=c("edge", "maSigPro", "Genes in common"), FDR_0.01=c("10", "135", "10"), FDR_0.1=c("61", "196", "55"))
formattable(DF)

```





### Compare the number of DE genes




### Compare the p-values and q-values of the genes that were statistically significant in both programs